Continuous developmental changes in word recognition across early childhood support early language learning

### 2024-11-17 5_longitudinal.qmd
bug fix and further development

---
title: "Longitudinal models"
format: html
---

Questions: 
(from Nilam)

1. Is the mean of the latent “language ability” factor scores different across 1st assessment, 2nd assessment, 3rd assessment, etc? This is basically an ANOVA on the latent factor scores – using “assessment number” as the categorical predictor. – an examination of “practice effects”.

Is the mean of the latent “language ability” factor scores increasing with age?
- This is basically a regression on the latent factor scores – and a growth model.
- In which case we also implicitly test research questions about the variances
- Is the (between-person) variance of the latent “language ability” factor scores increasing with age?

- A more nuanced version of the “practice effects" examination might use delta_t. 
- Is the mean of the latent “language ability” factor different across retest interval?
- This is basically a regression on the latent factor scores.

- In which case we also implicitly test research questions about the variances
- Is the (between-person) variance of the latent “language ability” factor scores increasing with age?

3. If we are interested in the covariances –
- Is the structure of the latent “language ability” factor changing with age?
- This again gets us into a multi-group SEM –
- Testing whether the factor loading structure differs across groups.
- This is most easily invoked with using categorical “time variable” – e.g, assessment number, age-group.
- However, it is also possible to do using a continuous time variable (e.g., age). In this case we basically use a moving window to obtain the covariances and then basically use age as a moderator of the factor loadings to examine the age x factor loading interaction.

# Descriptives

## Longitudinal data

@@

d_sub_long <- d_sub |>
  filter(subject_id %in% longitudinal$subject_id) |>
  group_by(subject_id) |>
  arrange(administration_id) |>
  mutate(admin_num = 1:n(), 
         delta_t = c(0,diff(age)))

ggplot(d_sub_long, 
       aes(x = delta_t, fill = as.factor(admin_num))) +
  geom_histogram() + 
  facet_wrap(~dataset_name)
```
## Datasets with CDIs

Longitudinal data with CDI is just FMW2013 right now.

```{r}
fmw <- filter(d_sub_long, dataset_name == "fmw_2013") |>
  mutate(t = ifelse(age < 23, "t1","t2")) |>
  group_by(subject_id, t) |>
  summarise(age = mean(age), 
            log_rt = mean(log_rt), 
            prod = mean(prod), 
            long_window_accuracy = mean(long_window_accuracy),
            delta_t = mean(delta_t)) |>
  pivot_wider(names_from = "t", 
              values_from = c("age","log_rt","prod","long_window_accuracy", "delta_t"))

summary(lm(prod_t2 ~ long_window_accuracy_t1 * log_rt_t1 + age_t2, 
           data = fmw))
```

## Without CDIs

```{r}
ggplot(d_sub_long, 
       aes(x = age, y = long_window_accuracy, group = subject_id, col = dataset_name)) +
  geom_line(alpha = .3, position = position_jitter(width = .1, height = 0)) + 
  geom_smooth(aes(group = 1)) + 
  scale_y_log10() + 
  geom_hline(lty = 2, yintercept = .5)

@@

```{r}
ggplot(filter(d_sub_long, n_trials_rt > 3),
       aes(x = age, y = log_rt, group = subject_id, col = dataset_name)) +
  geom_line(alpha = .3, position = position_jitter(width = .1, height = 0)) + 
  geom_smooth(aes(group = 1)) + 
  scale_y_log10() 
```
# Q1: age-related change in the latent score

Growth in latent language factor. This is the same as the previous factor analysis, but we only look within child and within dataset.

## 1 factor approach

This has the benefits of simplicity, even though we know that the 1F model is "wrong."

```{r}
fa_model <-  "F1  =~ rt + rt_var + long_window_accuracy + long_window_acc_var + prod + comp"
fit <- cfa(fa_model, d_sub_mat_s, std.lv=TRUE, missing='fiml')

d_sub$f1 <- lavPredict(fit, newdata = d_sub_mat_s)

@@

```{r}
ggplot(d_sub |>
         filter(subject_id %in% longitudinal$subject_id), 
       aes(x = age, y = f1, group = subject_id, col = dataset_name)) + 
  geom_point(alpha = .4) + 
  geom_line(alpha = .4) + 
  geom_smooth(aes(group = dataset_name), method = "lm")
```
Note that nesting age slopes doesn't converge. Could go Bayesian if we cared...

```{r}
long_mod <- lmer(f1 ~ age + (1 | subject_id) + (1 | dataset_name), 
                 data = d_sub)

summary(long_mod)
```

## 2 factor approach

Let's try looking at the speed/accuracy model across time.

```{r}
fa2_model <-  "accuracy =~ long_window_accuracy + long_window_acc_var + comp + prod
               speed =~ rt + rt_var"

fit2 <- cfa(fa2_model, d_sub_mat_s, std.lv=TRUE, missing='fiml')

preds <- predict(fit2)

d_sub$accuracy <- preds[,1]
d_sub$speed <- preds[,2]
```

Merge with longitudinal data.

```{r}
d_long <- d_sub |>
  filter(subject_id %in% longitudinal$subject_id) |>
  group_by(subject_id) |>
  arrange(administration_id) |>
  mutate(admin_num = 1:n(), 
         delta_t = c(0,diff(age)), 
         speed_0 = speed[1]) |>
  group_by(dataset_name) |>
  mutate(speed_0_high = speed_0 > median(speed_0, na.rm=TRUE))
```

Plot.

```{r}
ggplot(d_long |>
         pivot_longer(speed:accuracy, names_to = "variable", values_to = "value"), 
       aes(x = age, y = value, group = subject_id, col = dataset_name)) + 
  geom_point(alpha = .4) + 
  geom_line(alpha = .4) + 
  geom_smooth(aes(group = dataset_name), method = "lm") + 
  scale_x_log10() +
  facet_wrap(~variable) +
  theme(legend.position = "bottom") + 
  xlab("Age (months)") + 
  ylab("Latent score") 
```
Change in the latents over age.

```{r}
long_mod_acc <- lmer(accuracy ~ age + (1 | subject_id) + (1 | dataset_name), 
                 data = d_sub)
long_mod_speed <- lmer(speed ~ age + (1 | subject_id) + (1 | dataset_name), 
                 data = d_sub)

summary(long_mod_acc)
summary(long_mod_speed)
```

# Q2: relations between latents and other factors

Longitudinal model.

```{r}
lmod <- lmer(accuracy ~ age + speed_0 * delta_t + (1 | subject_id) + (1 | dataset_name), 
     data = d_long)

summary(lmod)
```

```{r}
ggplot(d_long, 
       aes(x = age, y = accuracy, group = subject_id, col = dataset_name, lty = speed_0_high)) + 
  geom_line(alpha = .4) + 
  geom_smooth(aes(group = speed_0_high), method = "lm") + 
  facet_wrap(~dataset_name, scales="free_x")
```

>>>>

---
title: "Longitudinal models"
format: 
  html:
    toc: true
execute: 
  cache: true
---

Notes from Nilam:
Look at this part of the latent growth tutorial #observed residual variances (constrained to equality)- math2 ~~ sigma2_u*math2

the two-wave model:
change in inter-individual differences
vs. differences in intra-individual change

This markdown documents attempts to do longitudinal models for the Peekbank data.

```{r}
library(here)
source(here("helper","common.R"))
library(lavaan)
library(tidySEM)

d_sub <- readRDS(here("cached_intermediates","1_d_sub.Rds")) |>
  group_by(subject_id) |>
  arrange(age) |>
  mutate(admin_num = 1:n(), 
         time_since_t0 = age - age[1],
         delta_t = c(0, diff(age)))

@@

d_sub_long <- d_sub |>
  filter(subject_id %in% longitudinal$subject_id)

ggplot(d_sub_long, 
       aes(x = time_since_t0, fill = dataset_name)) +
  geom_histogram(binwidth = 1) + 
  # facet_wrap(~dataset_name) + 
  xlab("Time since first test")
```

# Reliability

How to do this right? WRONG RIGHT NOW

```{r}
MONTH_CUTOFF <- 3

d_reliability <- d_sub_long |>
  filter(admin_num <= 2) |>
  pivot_wider(id_cols = c("subject_id","dataset_name"), 
              names_from = "admin_num",
              values_from = c("log_rt", "log_rt_var", "long_window_accuracy", 
                              "long_window_acc_var", "prod", "comp"))

dataset_reliabilities <- d_reliability |>
  group_by(dataset_name) |>
  summarise(rt = cor(log_rt_1, log_rt_2, use = "pairwise.complete.obs"),
            acc = cor(long_window_accuracy_1, 
                          long_window_accuracy_2, use = "pairwise.complete.obs")) |>
  filter(dataset_name != "fernald_totlot") 
```

# Longitudinal factor analysis

```{r}
d_sub_s <- d_sub |>
  ungroup() |>
  select(dataset_name, subject_id, administration_id, age, 
         time_since_t0, delta_t,
         log_rt, log_rt_var, 
         long_window_accuracy, long_window_acc_var, prod, comp, ) |>
  rename(acc = long_window_accuracy, 
         acc_sd = long_window_acc_var, 
         log_rt_sd = log_rt_var) |>
  ungroup() |>
  mutate(across(all_of(c("log_rt", "log_rt_sd", "acc", "acc_sd", 
                         "prod", "comp")), 
                       ~ age_scale(.x, age))) 
```

Check on amount of data.

# Two-time point

```{r}
d_sub_wide <- d_sub_s |> 
  mutate(t = cut(time_since_t0, breaks = c(0, 3, 6, 30), 
                 include.lowest = TRUE)) |>
  filter(t != "(6,30]") |>
  mutate(t = as.numeric(t)) |>
  group_by(subject_id, dataset_name, t) |>
  summarise(across(all_of(c("log_rt", "log_rt_sd", "acc",
                            "acc_sd", "prod", "comp")),
                   ~ mean(.x, na.rm=TRUE))) |> 
  mutate(across(all_of(c("log_rt", "log_rt_sd", "acc",
                         "acc_sd", "prod", "comp")),
                ~ ifelse(is.nan(.x), NA, .x))) |>
  pivot_wider(id_cols = c("subject_id","dataset_name"), 
              names_from = "t",
              values_from = c("log_rt", "log_rt_sd", "acc", 
                              "acc_sd", "prod", "comp"), 
              names_prefix = "t")

d_sub_wide |>
  ungroup() |>
  summarise(across(log_rt_t1:comp_t2, 
                   ~sum(!is.na(.x))))
```

```{r}
fa3_model_long <- "
# measurement model
vocab_t1 =~ 1*prod_t1 + s1*comp_t1
accuracy_t1 =~ 1*acc_t1 + s2*acc_sd_t1
speed_t1 =~ 1*log_rt_t1 + s3*log_rt_sd_t1

vocab_t2 =~ 1*prod_t2 + s1*comp_t1
accuracy_t2 =~ 1*acc_t2 + s2*acc_sd_t2
speed_t2 =~ 1*log_rt_t2 + s3*log_rt_sd_t2

# longitudinal relationships
vocab_t2 ~ vocab_t1 + speed_t1 + accuracy_t1
speed_t2 ~ vocab_t1 + speed_t1 + accuracy_t1
accuracy_t2 ~ vocab_t1 + speed_t1 + accuracy_t1

# factor covariances
vocab_t1 ~~ accuracy_t1
vocab_t1 ~~ speed_t1
accuracy_t1 ~~ speed_t1

vocab_t2 ~~ accuracy_t2
vocab_t2 ~~ speed_t2
accuracy_t2 ~~ speed_t2

# means for the latents
vocab_t1 ~ 1
vocab_t2 ~ 1
accuracy_t1 ~ 1
accuracy_t2 ~ 1
speed_t1 ~ 1
speed_t2 ~ 1

# residual variance for the latents
vocab_t1 ~~ NA*vocab_t1
vocab_t2 ~~ NA*vocab_t2
accuracy_t1 ~~ NA*accuracy_t1
accuracy_t2 ~~ NA*accuracy_t2
speed_t1 ~~ NA*speed_t1
speed_t2 ~~ NA*speed_t2
"

fit3_long <- sem(fa3_model_long, d_sub_wide, std.lv=TRUE, missing='fiml')

summary(fit3_long, fit.measures=TRUE, standardize=TRUE)

```
```{r}
layout_long = matrix(nrow=7, ncol = 7,
                     data = c("log_rt_sd_t1","log_rt_t1", "acc_t1", "acc_sd_t1", "prod_t1","comp_t1", NA,
                               NA, NA, NA, NA, NA, NA,  NA,
                              "speed_t1",NA,"accuracy_t1",NA,"vocab_t1",NA,  NA,
                              NA, NA, NA, NA, NA, NA,  NA,
                               NA,"speed_t2",NA,"accuracy_t2",NA,NA,"vocab_t2",
                               NA, NA, NA, NA, NA, NA,  NA,
                               NA,"log_rt_sd_t2","log_rt_t2", "acc_t2", "acc_sd_t2", "prod_t2", "comp_t2"), 
                     byrow = TRUE)
graph_sem(model = fit3_long, text_size = 2, layout = t(layout_long))
```

# Full longitudinal model

```{r}
d_sub_wide <- d_sub_s |> 
  mutate(t = cut(time_since_t0, breaks = c(0, 3, 6, 9, 12, 30), 
                 include.lowest = TRUE)) |>
  filter(t != "(12,30]") |>
  mutate(t = as.numeric(t)) |>
  group_by(subject_id, dataset_name, t) |>
  summarise(across(all_of(c("log_rt", "log_rt_sd", "acc",
                            "acc_sd", "prod", "comp")),
                   ~ mean(.x, na.rm=TRUE))) |> 
  mutate(across(all_of(c("log_rt", "log_rt_sd", "acc",
                         "acc_sd", "prod", "comp")),
                ~ ifelse(is.nan(.x), NA, .x))) |>
  pivot_wider(id_cols = c("subject_id","dataset_name"), 
              names_from = "t",
              values_from = c("log_rt", "log_rt_sd", "acc", 
                              "acc_sd", "prod", "comp"), 
              names_prefix = "t")

d_sub_wide |>
  ungroup() |>
  summarise(across(log_rt_t1:comp_t4, 
                   ~sum(!is.na(.x))))
```

```{r}
fa3_model_long <- "
# measurement model
vocab_t1 =~ 1*prod_t1 + s1*comp_t1
accuracy_t1 =~ 1*acc_t1 + s2*acc_sd_t1
speed_t1 =~ 1*log_rt_t1 + s3*log_rt_sd_t1

vocab_t2 =~ 1*prod_t2 
accuracy_t2 =~ 1*acc_t2 + s2*acc_sd_t2
speed_t2 =~ 1*log_rt_t2 + s3*log_rt_sd_t2

vocab_t3 =~ 1*prod_t3 
accuracy_t3 =~ 1*acc_t3 + s2*acc_sd_t3
speed_t3 =~ 1*log_rt_t3 + s3*log_rt_sd_t3

vocab_t4 =~ 1*prod_t4 
accuracy_t4 =~ 1*acc_t4 + s2*acc_sd_t4
speed_t4 =~ 1*log_rt_t4 + s3*log_rt_sd_t4

# factor covariances
vocab_t1 ~~ accuracy_t1
vocab_t1 ~~ speed_t1
accuracy_t1 ~~ speed_t1

vocab_t2 ~~ accuracy_t2
vocab_t2 ~~ speed_t2
accuracy_t2 ~~ speed_t2

vocab_t3 ~~ accuracy_t3
vocab_t3 ~~ speed_t3
accuracy_t3 ~~ speed_t3

vocab_t4 ~~ accuracy_t4
vocab_t4 ~~ speed_t4
accuracy_t4 ~~ speed_t4

# means for the latents
vocab_t1 ~ 1
vocab_t2 ~ 1
vocab_t3 ~ 1
vocab_t4 ~ 1
accuracy_t1 ~ 1
accuracy_t2 ~ 1
accuracy_t3 ~ 1
accuracy_t4 ~ 1
speed_t1 ~ 1
speed_t2 ~ 1
speed_t3 ~ 1
speed_t4 ~ 1

# regressions
accuracy_intercept =~ 1*accuracy_t1 + 1*accuracy_t2 + 1*accuracy_t3 + 1*accuracy_t4 
accuracy_slope =~ 1*accuracy_t1 + 2*accuracy_t2 + 3*accuracy_t3 + 4*accuracy_t4
speed_intercept =~ 1*speed_t1 + 1*speed_t2 + 1*speed_t3 + 1*speed_t4 
speed_slope =~ 1*speed_t1 + 2*speed_t2 + 3*speed_t3 + 4*speed_t4
vocab_intercept =~ 1*vocab_t1 + 1*vocab_t2 + 1*vocab_t3 + 1*vocab_t4 
vocab_slope =~ 1*vocab_t1 + 2*vocab_t2 + 3*vocab_t3 + 4*vocab_t4

# parameters for the regressions
vocab_intercept ~ 1
vocab_slope ~ 1
accuracy_intercept ~ 1
accuracy_slope ~ 1
speed_intercept ~ 1
speed_slope ~ 1

# # residual variance for the latents
# vocab_t1 ~~ NA*vocab_t1
# vocab_t2 ~~ NA*vocab_t2
# accuracy_t1 ~~ NA*accuracy_t1
# accuracy_t2 ~~ NA*accuracy_t2
# speed_t1 ~~ NA*speed_t1
# speed_t2 ~~ NA*speed_t2
"

fit3_long <- sem(fa3_model_long, d_sub_wide, std.lv=TRUE, missing='fiml')

summary(fit3_long, fit.measures=TRUE, standardize=TRUE)

@@

```{r}
layout_dt = matrix(nrow=11, ncol = 7,
                   data = c("prod_t1", "prod_t2", "prod_t3", "prod_t4", "vocab_intercept", NA, NA,
                            NA, NA, NA, NA, NA, NA, NA,
                            "comp_t1", NA, NA, NA, "vocab_slope", NA,NA,
                            NA, NA, NA, NA, NA, NA, NA,
                            "accuracy_t1", "accuracy_t2", "accuracy_t3", "accuracy_t4", "accuracy_intercept", NA, NA,
                            NA, NA, NA, NA, NA, NA, NA,
                            "acc_sd_t1", "acc_sd_t2", "acc_sd_t3", "accuracy_sd_t4", "accuracy_slope", NA, NA,
                            NA, NA, NA, NA, NA, NA, NA,
                            "log_rt_t1", "log_rt_t2", "log_rt_t3", "log_rt_t4", "speed_int", NA, NA,
                            NA, NA, NA, NA, NA, NA, NA,
                            "log_rt_sd_t1", "log_rt_sd_t2", "log_rt_sd_t3", "log_rt_sd_t4", "speed_slope", NA, NA),
                   byrow = TRUE)

graph_sem(model = fit3_long, text_size = 3, layout = layout_dt)

@@

### 2024-11-17 paper/paper.Rmd
bug fix and further development

```{r loading, echo=FALSE, message=FALSE, warning=FALSE}
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(psych))
suppressPackageStartupMessages(library(lmerTest))
suppressPackageStartupMessages(library(lavaan))
suppressPackageStartupMessages(library(tidySEM))
suppressPackageStartupMessages(library(xtable))
# suppressPackageStartupMessages(library(kableExtra))
suppressPackageStartupMessages(library(papaja))

source(here("helper","common.R"))
d_trial <- readRDS(here("cached_intermediates","1_d_trial.Rds")) |>
  mutate(log_age = log(age))
d_sub <- readRDS(here("cached_intermediates","1_d_sub.Rds"))

@@

These criteria yielded `r length(unique(d_sub$dataset_name))` datasets, including `r length(unique(d_sub$subject_id))` children and `r length(unique(d_sub$administration_id))` administrations of the LWL procedure (some datasets were longitudinal or involved multiple closely-spaced testing sessions). Table \ref{tab:datasets} shows the characteristics of individual datasets. The size of the combined dataset, the unified data processing pipeline, and the fact that individual studies used very similar implementations of the LWL experimental paradigm, all allow us to take a more detailed look at early language processing than has previously been possible.

@@

We were first interested in developmental changes in speed and accuracy. We computed both reaction times and accuracies following standard practices in the literature [@fernald2008]. Because there is no consensus about the length of time windows for the computation of accuracy, we considered both a shorter window (from 200 -- 2200 ms after noun onset) and a longer window (from 200 -- 4000 ms).

Our first question was about the functional form of the relationships between age, speed, and accuracy. To investigate this question, we fit linear mixed effects models predicting accuracy and reaction time on each trial across the full dataset with random slopes of age by study and random intercepts by participant. We compared models that included both long and short accuracy windows, as well as logarithmic and linear effects of age, and logarithmic and linear transformations of reaction time. The best fitting model of accuracy predicted long window accuracy as a function of the logarithm of age,^[This longer window yielded overall higher cross-trial reliability as well, and so we report long window results in the main text. All findings reported here hold for both window sizes, however.] and the best fitting model of speed predicted log reaction time as a function of log age as well (See Supplemental Tables 1 and 2 for model fit statistics).

@@

## Speed and accuracy covary with vocabulary {.unnumbered}

@@

We fit a series of factor analytic models to explore the dimensionality of the data. Initial exploratory factor analysis using paralell analysis suggested that three factors explained substantial variance in the data. Due to missingness of data, we used confirmatory factory analysis with full information maximum likelihood to find the best set of loadings. The best fitting model was a three-factor model with factors for speed (reaction time and reaction time variability), accuracy (accuracy and accuracy variability), and vocabulary (comprehension and production from the CDI). Fit statistics for this model were generally good (Confirmatory fit index: `r round(fit_stats$fit["cfi"], 3)`, RMSE: `r round(fit_stats$fit["rmsea"], 3)`, RMSE $p$-value: `r round(fit_stats$fit["rmsea.pvalue"], 3)`).

```{r age_regression, fig.env="figure*", fig.cap="Structural equation model showing the three-factor factor analysis with a regression of each latent variable on the logarithm of age. Squares show  Solid lines show regression coefficients and factor loadings, and dotted lines show covariances (reflecting age-residualized correlations). Abbreviations: acc = accuracy, prod = production vocabulary, comp = comprehension vocabulary.", fig.height = 3}

@@

d_long <- d_sub |>
  filter(subject_id %in% longitudinal$subject_id) |>
  group_by(subject_id) |>
  arrange(administration_id) |>
  mutate(admin_num = 1:n(), 
         delta_t = c(0,diff(age)))

@@

```{r prep_long_data}
d_sub_long <- d_sub |>
  mutate(across(all_of(c("log_rt", "log_rt_var", "long_window_accuracy",
                         "long_window_acc_var", "prod", "comp")),
                ~ age_scale(.x,age))) |>
  group_by(subject_id) |>
  arrange(administration_id) |>
  mutate(admin_num = 1:n(), 
         delta_t = c(0, diff(age)))

d_sub_long_t <- d_sub_long |> 
  mutate(t = ifelse(delta_t > MONTH_CUTOFF, "t2", "t1")) |>
  rename(acc = long_window_accuracy,
         acc_sd = long_window_acc_var,
         log_rt_sd = log_rt_var) |>
  group_by(subject_id, dataset_name, t) |>
  summarise(across(all_of(c("log_rt", "log_rt_sd", "acc",
                            "acc_sd", "prod", "comp")),
                   ~ mean(.x, na.rm=TRUE))) |> 
  mutate(across(all_of(c("log_rt", "log_rt_sd", "acc",
                         "acc_sd", "prod", "comp")),
                ~ ifelse(is.nan(.x), NA, .x)))

d_sub_wide <- d_sub_long_t |>
  pivot_wider(id_cols = c("subject_id","dataset_name"), 
              names_from = "t",
              values_from = c("log_rt", "log_rt_sd", "acc", 
                              "acc_sd", "prod", "comp"))
```

@@

@@

Given this relatively low reliability in our longitudinal datasets (which also tended to be from younger children), we chose to measure longitudinal relationships between LWL and vocabulary via a longitudinal latent factor model. We fit the same three-factor model, but this time averaging longitudinal administrations after splitting them into two categories: those datapoints within a two-month time period (which we designate as $t_1$) and those datapoints between three and six months from the initial measurement ($t_2$). To ensure measurement invariance, we coupled the factor loadings across the two time-points. Next, to investigate coupling between the different measures, we fit regressions linking the two time points.

The fitted longitudinal model is shown in \ref{fig:longitudinal}. Overall fit statistics were generally good (Confirmatory fit index: `r round(fit_stats$fit["cfi"], 3)`, RMSE: `r round(fit3_long_summary$fit["rmsea"], 3)`, RMSE $p$-value: `r round(fit3_long_summary$fit["rmsea.pvalue"], 3)`). Each latent factor was significantly coupled across time, again indicating moderate longitudinal stability of these abilities.

```{r longitudinal, fig.env="figure*", height = 3.5, fig.cap = "Coupled change model."}
layout_long = matrix(nrow=7, ncol = 7,
                     data = c("log_rt_sd_t1","log_rt_t1", "acc_t1", "acc_sd_t1", "prod_t1","comp_t1", NA,
                               NA, NA, NA, NA, NA, NA,  NA,
                              "speed_t1",NA,"accuracy_t1",NA,"vocab_t1",NA,  NA,
                              NA, NA, NA, NA, NA, NA,  NA,
                               NA,"speed_t2",NA,"accuracy_t2",NA,NA,"vocab_t2",
                               NA, NA, NA, NA, NA, NA,  NA,
                               NA,"log_rt_sd_t2","log_rt_t2", "acc_t2", "acc_sd_t2", "prod_t2","comp_t2"), 
                     byrow = TRUE)
graph_sem(model = fit3_long, text_size = 2, layout = t(layout_long))
```

@@

Critically, LWL processing speed at time 1 predicted vocabulary at time 2 ($\beta=$ `r round(fit3_long_summary$pe[14,"est"],3)`, SE = `r round(fit3_long_summary$pe[21,"se"],3)`, $p =$ `r round(fit3_long_summary$pe[14,"pvalue"],3)`). This model significantly improved fit compared with a model that set this coupling parameter to zero but was otherwise unchanged ($\chi^2 = `r round(speed_vocab_anova$c2d[2], 2)`$, $p =  `r round(speed_vocab_anova$p[2], 4)`$). These findings are consistent with the claim that faster processing is related to longitudinal growth in vocabulary [@fernald2012,@Weisleder2013].

@@

We used a new database aggregating results across many individual developmental eye-tracking experiments to investigate the developmental course of word recognition and its relationship with early language learning. We found continuous developmental changes across all measures, consistent with a characterization of word recognition as a skill. Speed and accuracy both improved asymptoticaly, with evidence that recognition speed showed the log-log relationship associated with the power law of practice, that is, gradually converging on mature levels of processing efficiency. Further, older children showed less variability in accuracy and reaction time across trials, indicating more robust execution of this skilled behavior.

>>>>

```{r loading, echo=FALSE, message=FALSE, warning=FALSE}
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(psych))
suppressPackageStartupMessages(library(lmerTest))
suppressPackageStartupMessages(library(lavaan))
suppressPackageStartupMessages(library(tidySEM))
suppressPackageStartupMessages(library(xtable))
suppressPackageStartupMessages(library(papaja))

source(here("helper","common.R"))

d_trial <- readRDS(here("cached_intermediates","1_d_trial.Rds")) |>
  mutate(log_age = log(age))
d_sub <- readRDS(here("cached_intermediates","1_d_sub.Rds")) |>
  group_by(subject_id) |>
  arrange(age) |>
  mutate(admin_num = 1:n(), 
         time_since_t0 = age - age[1],
         delta_t = c(0, diff(age)))

@@

These criteria yielded `r length(unique(d_sub$dataset_name))` datasets, including `r length(unique(d_sub$subject_id))` children and `r length(unique(d_sub$administration_id))` administrations of the LWL procedure (some datasets were longitudinal or involved multiple closely-spaced testing sessions). Table \ref{tab:datasets} shows the characteristics of individual datasets (see also Figures S1 and S2). The size of the combined dataset, the unified data processing pipeline, and the fact that individual studies used very similar implementations of the LWL experimental paradigm, all allowed us to make a more detailed exploration of the development of word recognition than has previously been possible.

@@

We were first interested in developmental changes in speed and accuracy. We computed both reaction times and accuracies following standard practices in the literature [@fernald2008]. Because there is no consensus about the length of time windows for the computation of accuracy, we considered both a shorter window (from 200 -- 2200 ms after noun onset) and a longer window (from 200 -- 4000 ms). 
Our first question was about the functional form of the relationships between age, speed, and accuracy (see Table S-XYZ for raw correlations between variables). To investigate this question, we fit linear mixed effects models predicting accuracy and reaction time on each trial across the full dataset with random slopes of age by study and random intercepts by participant. We compared models that included both long and short accuracy windows, as well as logarithmic and linear effects of age, and logarithmic and linear transformations of reaction time. The best fitting model of accuracy predicted long window accuracy as a function of the logarithm of age,^[This longer window yielded overall higher cross-trial reliability as well, and so we report long window results in the main text. All findings reported here hold for both window sizes, however (see SI: Short Window Analyses).] and the best fitting model of speed predicted log reaction time as a function of log age as well (See Tables S1 and S2 for model fit statistics).

@@

## Speed and accuracy relate to vocabulary size {.unnumbered}

@@

We fit a series of factor analytic models to explore the dimensionality of the data. Initial exploratory factor analysis using paralell analysis suggested that three factors explained substantial variance in the data (see SI: Factor Analysis). Due to missingness of data, we used confirmatory factory analysis with full information maximum likelihood to find the best set of loadings. The best fitting model was a three-factor model with factors for speed (reaction time and reaction time variability), accuracy (accuracy and accuracy variability), and vocabulary (comprehension and production from the CDI). Fit statistics for this model were generally good (Confirmatory fit index: `r round(fit_stats$fit["cfi"], 3)`, RMSE: `r round(fit_stats$fit["rmsea"], 3)`)`; see SI: Alternative Factor Structures).

```{r age_regression, fig.env="figure*", fig.cap="Structural equation model showing the three-factor factor analysis with a regression of each latent variable on the logarithm of age. Observed variables are notated as squares and latent variables are notated as circles Factor loadings and regression coefficients are shown with straight, solid lines; covariances are shown with dashed lines; residual variances are shown as solid circular connections. Stars show conventional levels of statistical significance, e.g. * indicates p < .05, ** indicates p < .01, and *** indicates p < .001. Covariances reflect age-residualized correlations between variables. Abbreviations: acc = accuracy, prod = production vocabulary, comp = comprehension vocabulary.", fig.height = 3}

@@

d_long <- d_sub |>
  filter(subject_id %in% longitudinal$subject_id) |>
  group_by(subject_id) |>
  arrange(age) |>
  mutate(admin_num = 1:n(), 
         delta_t = age - age[1])

@@

```{r prep_long_data}
# rename and scale variables
d_sub_s <- d_sub |>
  ungroup() |>
  select(dataset_name, subject_id, administration_id, age, 
         time_since_t0, delta_t,
         log_rt, log_rt_var, 
         long_window_accuracy, long_window_acc_var, prod, comp, ) |>
  rename(acc = long_window_accuracy, 
         acc_sd = long_window_acc_var, 
         log_rt_sd = log_rt_var) |>
  ungroup() |>
  mutate(across(all_of(c("log_rt", "log_rt_sd", "acc", "acc_sd", 
                         "prod", "comp")), 
                       ~ age_scale(.x, age)))

d_sub_wide <- d_sub_s |> 
  mutate(t = cut(time_since_t0, breaks = c(0, 6, 12, 30), 
                 include.lowest = TRUE)) |>
  filter(t != "(12,30]") |>
  mutate(t = as.numeric(t)) |>
  group_by(subject_id, dataset_name, t) |>
  summarise(across(all_of(c("log_rt", "log_rt_sd", "acc",
                            "acc_sd", "prod", "comp")),
                   ~ mean(.x, na.rm=TRUE))) |> 
  mutate(across(all_of(c("log_rt", "log_rt_sd", "acc",
                         "acc_sd", "prod", "comp")),
                ~ ifelse(is.nan(.x), NA, .x))) |>
  pivot_wider(id_cols = c("subject_id","dataset_name"), 
              names_from = "t",
              values_from = c("log_rt", "log_rt_sd", "acc", 
                              "acc_sd", "prod", "comp"), 
              names_prefix = "t")

```

@@

vocab_t2 =~ 1*prod_t2 
accuracy_t2 =~ 1*acc_t2 + s2*acc_sd_t2
speed_t2 =~ 1*log_rt_t2 + s3*log_rt_sd_t2

@@

Given this relatively low reliability in our longitudinal datasets (which also tended to be from younger children), we chose to measure longitudinal relationships between LWL and vocabulary via a longitudinal latent factor model. We fit the same three-factor model, but this time averaging longitudinal administrations after splitting them into two categories: those datapoints within a two-month time period (which we designate as $t_1$) and those datapoints between three and six months from the initial measurement ($t_2$) (see SI: Sensitivity Analysis for Longitudinal Models). To ensure measurement invariance, we coupled the factor loadings across the two time-points. Next, to investigate coupling between the different measures, we fit regressions linking the two time points.

The fitted longitudinal model is shown in Figure \ref{fig:longitudinal}. Overall fit statistics were generally good (Confirmatory fit index: `r round(fit_stats$fit["cfi"], 3)`, RMSE: `r round(fit3_long_summary$fit["rmsea"], 3)`, RMSE $p$-value: `r round(fit3_long_summary$fit["rmsea.pvalue"], 3)`). Each latent factor was significantly coupled across time, again indicating moderate longitudinal stability of these abilities.

```{r longitudinal, fig.env="figure*", height = 3.5, fig.cap = "Structural equation model showing longitudinal couplings between latent factors. Variables marked t1 are measures from two or fewer months from study initiation; variables marked t2 are measures from 3--6 months after study initiation. Factor loadings are constrained to be equivalent between t1 and t2. Plotting conventions are as in Figure 3."}
layout_long = matrix(nrow=7, ncol = 7,
                     data = c("log_rt_sd_t1","log_rt_t1", "acc_t1", "acc_sd_t1", "prod_t1","comp_t1", NA,
                               NA, NA, NA, NA, NA, NA,  NA,
                              "speed_t1",NA,"accuracy_t1",NA,"vocab_t1",NA,  NA,
                              NA, NA, NA, NA, NA, NA,  NA,
                               NA,"speed_t2",NA,"accuracy_t2",NA,NA,"vocab_t2",
                               NA, NA, NA, NA, NA, NA,  NA,
                               NA,"log_rt_sd_t2","log_rt_t2", "acc_t2", "acc_sd_t2", NA, "prod_t2"), 
                     byrow = TRUE)
graph_sem(model = fit3_long, text_size = 2, layout = t(layout_long))
```

@@

Critically, LWL processing speed at time 1 predicted vocabulary at time 2 ($\beta=$ `r round(fit3_long_summary$pe[14,"est"],3)`, SE = `r round(fit3_long_summary$pe[21,"se"],3)`, $p =$ `r round(fit3_long_summary$pe[14,"pvalue"],3)`). This model significantly improved fit compared with a model that set this coupling parameter to zero but was otherwise unchanged ($\chi^2 = `r round(speed_vocab_anova$c2d[2], 2)`$, $p =  `r round(speed_vocab_anova$p[2], 4)`$). These findings are consistent with the claim that faster processing is related to longitudinal growth in vocabulary [@fernald2006;@Weisleder2013].

@@

How does word recognition change across early childhood, and how does it relate to language learning? We investigated these questions using a new dataset of developmental eye-tracking measurements. We found continuous developmental changes from ages 1 -- 6 years. Speed and accuracy both improved asymptotically, with evidence that recognition speed showed the log-log relationship associated with the power law of practice, that is, gradually converging on mature levels of processing efficiency. Further, older children showed less variability in accuracy and reaction time across trials, indicating more robust execution of this skilled behavior. Speed and accuracy were both related to vocabulary size concurrently, and processing speed was also related to later vocabulary growth.

By aggregating data from many pre-existing studies, we were able to overcome the limitations of prior investigations, which typically had samples sizes at least an order of magnitude smaller than ours. The presence of multiple studies with longitudinal data,

These findings are consistent with theories that posit that language learning is a process of skill acquisition, in which children become skillful at quickly converting ephemeral signals into meaning [@christiansen2016].
